Mechanistic Modeling and Analysis of the Mammalian Unfolded Protein Response

### 2016-06-20 scientific-reports-title/Paper_v2.tex
Working my way through the results

\begin{abstract}
%Cellular stressors routinely lead to malfunctions in the folding state of critical proteins, thereby playing a vital role in disease states like cancer, diabetes and cardiovascular ailments. Newly synthesized proteins must fold and assemble into unique three-dimensional structures, in order to become functionally active. Cells monitor protein folding by an inbuilt quality-control system involving both the Endoplasmic Reticulum (ER) and the Golgi apparatus. Incorrectly folded proteins are tagged for degradation via ER associated degradation (ERAD) or sent back through a refolding cycle. However, accumulation of incorrectly folded proteins can trigger a cascade of events, termed the Unfolded Protein Response (UPR). UPR leads to re-establishing the cellular homeostasis. %This is done by increased transcription of chaperones, reduced translation and cellular apoptosis.
We developed a family of mechanistic models of the eukaryotic Unfolded Protein Response (UPR), which was composed of a system of ordinary differential equations.
The objective of this study was to assemble a series of molecular modules describing different aspects of UPR and subsequently analyze these models for fragility and robustness.
Kinetic parameters for these models were estimated by comparing simulations with experimental data. Using POETs and a cross-validation scheme, we developed an ensemble of models, consistent with literature data.
Model analysis highlighted the presence of a sequential order for firing of the ER stress transducers in UPR. The counteracting effects of the ER stress transducers converge at feedback regulation of the molecular chaperone BiP.
The three main regulators of feedback, ATF4, cleaved ATF6 and XBP1s share the load ensuring the phased response of UPR. However these regulators add on to the overall fragility of the system and allow scope for manipulation of UPR as seen by sensitivity/robustness analysis.
Downstream effects of UPR include a balance between cell survival and cell death based on the magnitude of the stress.
Model analysis suggested that the cell-death axis was relatively robust owing to redundant routes e.g., APAF-1 dependent and APAF-1 independent routes of apoptosis.
However the cell-survival axis was relatively susceptible to perturbations owing to multiple levels of regulation, thereby highlighting importance of key proteins like CHOP.
\end{abstract}
\begin{document}

@@

\subsection*{Formulation of the UPR network architecture.}
The UPR network described the ER folding cycle, ER-associated degradation (ERAD), ER-stress transducer (PERK, IRE1 and ATF6) signaling cascades and stress-induced caspase activation (Fig. \ref{fg:fig1}).
The network consisted of 636 protein or mRNA species interconnected by 1090 interactions (Fig. \ref{fg:fig1} Inset).
Connectivity was formulated from a comprehensive review of the primary literature  \cite{naidoo2009er, ron2002translational, kaufman2002unfolded,ellgaard2003qce,Fonseca:2009fk,Schroder:2005vn,gotoh:hdc,mccullough2001gsc, belmont2008cga,Hetz:2009qc,Urano:2000uq,szegezdi2006mediators}, and from on-line databases; String-8 \cite{Jensen:2009fk}, NetworKIN \cite{Linding:2007fk} and TRANSFAC.
Model connectivity was not specific to a single cell-line; rather, it was a canonical representation of the pathways involved in monitoring and controlling the folding capacity of a generic well-mixed ER compartment.
UPR induction was modeled as the release of BiP from the ER stress transducers, PERK, IRE1$\alpha$, and ATF6 leading initially to adaptation of the folding cycle and then, subsequently, to alarm and apoptosis.
The adaption phase of UPR was marked by general translation attenuation, selective transcriptional programs for key species like bZIP transcription factor ATF4 \cite{lu2004tra}, cellular inhibitor of apoptosis (cIAP) \cite{hamanaka2008pdr}, molecular chaperones e.g., BiP \cite{harding2003isr}, and enhanced clearance of accumulated proteins via ERAD.
The alarm and apoptosis phases were mediated by the induction of CHOP \cite{ron1992cnd}, regulation of Bcl2, Bcl2-antagonist of cell death (BAD) \cite {wang1999cia} and (TNF) receptor associated factor 2 (TRAF2) \cite{lei2003jpb,putcha2003jmb, yamamoto1999bpa, szegezdi2006mediators} activation (Fig. \ref{fg:fig1}).
Model connectivity is available in SBML format from the supplemental materials. Complete details of the interactions in the model are also enlisted in the supplementary materials.

%\clearpage
% \subsection*{Estimating a population of canonical UPR models using POETs.}
\subsection*{The population of UPR models recapitulated adaptation, alarm, and apoptotic events across multiple cell-lines and timescales.}
Despite a significant identifiability challenge, the multiobjective POETs algorithm generated a predictive UPR model population.
The three phases of UPR were modeled using mass action kinetics within an ordinary differential equation (ODE) framework.
While ODEs and mass-action kinetics are common methods of modeling biological pathways \cite{Schoeberl:2002rv, Chen:2009uq, Helmy:2009kx}, this modeling strategy resulted in a large number of unknown model parameters.
These parameters were not uniquely identifiable (data not shown).
Instead, we estimated an experimentally constrained population of parameters using multiobjective optimization.
A population of the 1726 unknown model parameters (1090 kinetic parameters and 636 initial conditions) was estimated from 33 dynamic and steady state data-sets taken from literature (Table \ref{obj-table}).
The residual between model simulations and each of the experimental constraints was simultaneously minimized using the multiobjective POETs algorithm \cite{Song:2010ij}.
A leave-eight-out cross-validation strategy was used to independently estimate the training and prediction error over the 33 data sets; we estimated four different model families, where eight of the 33 objectives were reserved for validation and 25 were used for model training of each family.
Further details on model and training in supplementary materials.
Thus, each model family was trained and validated on different experimental data.
Starting from an initial best-fit initial parameter set (nominal set), more than 25,000 probable models were estimated by POETs from which we selected N = 100 models (25 from each training family) with a Pareto rank of one or less (from approximately 1200 possible choices) for further study.
The nominal, training (75 models), and prediction (25 models) errors were calculated for each objective (Table \ref{obj-table}).
Models used for prediction error calculations for a particular objective were \emph{not} trained on that objective.
The prediction likelihood was statistically significantly better for 31 of the 33 objective functions at a 95\% confidence level, compared with random parameter sets generated from the nominal set (Table \ref{obj-table}).
POETs generated model families that predicted approximately 94\% of the objective functions with a significantly higher likelihood than a random control.
However, the specific value of any given parameter was likely not well described (data not shown).
The coefficient of variation (CV) for the model parameters ranged from 0.5 - 1.6, where approximately 65\% of the parameters were constrained with a CV $\leq$ 1.0 (supplementary materials Fig. S\ref{fg:Para_S1}).
The most constrained parameters involved a wide-array of functions e.g., regulation of PERK, eIF2$\alpha$, ATF4, Calcineurin, BiP, CHOP and ATF6 signaling.
However, the least constrained parameters involved JNK and apoptosis interactions.

POETs identified Pareto fronts between several objectives, e.g., O13$\times$O14, O25$\times$O29, O11$\times$O29, and O27$\times$O2, in the training data  (Fig. \ref{fg:objective_plot}).
Strong Pareto fronts suggested an inability to simultaneously model different aspects of the training data.
However, fronts could also result from experimental artifacts, e.g., variation between cell-lines, time-scale differences, or from functional relationships in the data.
Globally, adaptation and alarm phase training constraints conflicted with those involving apoptosis.
For example, objectives involving caspase-7 or caspase-9 activity conflicted with phosphorylated eIF2$\alpha$ levels.
Phosphorylation of eIF2$\alpha$ by activated PERK attenuates translation, which decreases the ER folding load.
Thus, eIF2$\alpha$ phosphorylation is a key early adaptive event in UPR. On the other hand, caspase-9 is a stress-induced death marker activated only after UPR has failed to restore folding homeostasis.
Conflicts between these early and late phase markers suggested the UPR time-scale was perhaps cell-line or perturbation dependent.
Similarly, negative feedback may lead to conflicting objectives as well (Fig. \ref{fg:objective_plot} and supplementary information).

%For example, XBP1 mRNA measurements (O14) conflicted with CHOP protein measurements (O13), even though these data-sets were taken from the same study and were collected in the same cell-line. XBP1 splicing increased BiP levels, which in turn reduced CHOP protein levels, hence the trade-off. Lastly, in addition to fronts, we also observed strong correlation between objectives. For example, models that performed well for the CHOP protein (O11), also performed well against Procaspase-12 (O22) measurements, even though these were not in the same cell-line or from the same study. Both CHOP and Procaspase-12 are downstream of the IRE1/TRAF2/JNK signaling cascade, so these errors were directly correlated.
%\clearpage
% \subsection*{The population of UPR models recapitulated adaptation, alarm, and apoptotic events across multiple cell-lines and timescales.}
%The counteracting effects are further substantiated in simulated knockout studies, wherein knockout of one ER stress transducer led to enhancement of the other branches of UPR (supplementary materials Fig. S\ref{fg:Flux_Supp_Fig1}A-C).
%We compared the ability of the system to recuperate upon reversal of unfolded protein loads from the alarm phase (P4, Fig. \ref{fg:proof_of_concept_simulation} ) and the apoptosis phase (P2, Fig. \ref{fg:proof_of_concept_simulation} ). If we reduced the protein load after the cell was committed to apoptosis (as in P3), we found that the cell continued to function similar to the UPR state even upon protein load reduction after 25 hrs (Fig. \ref{fg:cross_plot_flux}C).

We trained the UPR ensemble using data from UPR initiation events (e.g., PERK activation) and the downstream activation of proteins involved in apoptosis (BiP, caspase-3 and caspase-7).
ER stress induced by exposure to thapsigargin (Tg), a non-competitive inhibitor of SERCA Ca$^{2+}$ transporters, leads to the dissociation of BiP from the ER-stress transducers.
For these initial proof-of-concept simulations, we assumed that the action of Tg and other stress-inducing agents such as dithiothreitol (DTT) was identical, i.e., induction of BiP dissociation.
Within 20 minutes after Tg exposure, PERK was activated and transmitted adaptation signals downstream.
The population of UPR models recapitulated the timescale of PERK phosphorylation (Fig. \ref{fg:simulation_comparison}A) as well as its downstream signaling activity, for example, the phosphorylation of eIF2$\alpha$ (Fig. \ref{fg:simulation_comparison}H).
The nuclear fraction of ATF4 increased from approximately zero (untreated cells) to a maximum value 4 hrs after Tg exposure.
While the model ensemble generally predicted the correct trend, there was significant error in the early time points for ATF4 (Fig. \ref{fg:simulation_comparison}G).
The phosphorylation of eIF2$\alpha$ by PERK is required for ATF4 activation. Interestingly, when we compared model simulations of p-eIF2$\alpha$ levels following Tg (1$\mu$M) exposure in mouse embryonic fibroblasts (MEFs) with measurements (O15 \cite{yamamoto2007tim}), the model correctly captured the appropriate behavior.
To test the functionality of the ATF6 branch of the UPR model, we compared simulations with measurements of cleaved ATF6 in tunicamycin-treated MEFs \cite{lee2002imu}.
ER stress is known to lead to the release of BiP from ATF6.
Cleaved ATF6 is then translocated to the nucleus where it up-regulates gene expression \cite{silver1999mtf,hai1989tfa}.
Simulations of cleaved ATF6 levels following UPR initiation were consistent with measurements (Fig. \ref{fg:simulation_comparison}C).
The formation of p-PERK or p-IRE1 initiates a complex series of events that operate on both short and long time-scales.
Signals from the ER stress transducers converge downstream to regulate BiP transcription \cite{malhotra2007era, rao2004mpe,kokame2001iei, yoshida2000aap}.
Model ensemble recapitulated correct trends of BiP mRNA with maximum levels $\sim$ 8 hrs, similar to what was seen in experiments done on HEK293 cells (Fig. \ref{fg:simulation_comparison}D) \cite{lin2007ire1}.
One of the long-term outcomes of PERK/IRE1 activation is apoptotic cell death. In the proof-of-concept ER-model the link between UPR and apoptosis occurred through the action of eIF2$\alpha$, the dual role of the ATF4 transcription factor and caspase-12 activation by IRE1-TRAF2 signaling axis.
We constrained model parameters associated with the activation of the cell-death program using measurements of pro/caspase-7 levels, pro/caspase-9 levels, pro/caspase-3 levels, pro/caspase-12 levels and PARP cleavage mediated by executioner caspases following treatment with 0.5$\mu$M Tg \cite{rao2002cer}.
These experiments were performed in Sak2 cells that lack Apaf-1 protein expression \cite{rao2002cer}.
Thus, the data allowed us to include a non-Apaf-1 mediated stress-induced caspase activation pathway into the model.
The population of models recapitulated caspase-3 (Fig. \ref{fg:simulation_comparison}K) and caspase-9 (Fig. \ref{fg:simulation_comparison}J), as well as cleaved Parp levels (Fig. \ref{fg:simulation_comparison}L) following exposure to ER stress-inducers.
Interestingly, while PERK activation occurred on the timescale of minutes, initiator and executioner caspase activation occurred over 36 hrs.
Thus, the population of UPR models captured complex signaling events occurring across multiple timescales.

\subsection*{Signal flow analysis of UPR highlighted modes of crosstalk and redundancy in BiP and Bcl2 regulation.}
Traditionally, it has been hypothesized that there is a sequential order for firing of the ER stress transducers in UPR.
The PERK and ATF6 branches are thought to be activated before IRE1 \cite{szegezdi2006mediators} and largely promote ER adaptation to misfolding, while IRE1 is the last one to fire as it has a dual role, transmitting both survival and pro-apoptotic signals.
However, our modeling analysis suggests that these three branches fire simultaneously with varying rates and that the state of the cell in terms of adaptation, alarm, or apoptosis is a result of counteracting effects of these three prongs of UPR signaling (Fig. \ref{fg:proof_of_concept_simulation} and supplementary materials Fig. S\ref{fg:Flux_Supp_Fig1}).
UPR induction in the model was controlled by manipulation of the generation rate of unfolded or misfolded protein (qP) in the ER compartment. Upon UPR induction, initially ($\leq$ 1hr) the response is damped marking the adaptation phase of UPR.
Adaptation is followed by an increase in the activity of the IRE1$\alpha$, PERK, and ATF6 cascades at $\sim$ 1 hr marking the onset of the alarm phase, which leads to a final steady state $\sim$ 8-10 hrs marking the onset of the commitment or apoptosis phase of UPR (Fig. \ref{fg:proof_of_concept_simulation}).
This time frame was consistent with the maximum levels of BiP mRNA upon UPR induction \cite{lin2007ire1}. Our analysis was substantiated further by looking at the fluxes at different phases of UPR induction. As compared to P1 (No-UPR Steady State Fig. \ref{fg:proof_of_concept_simulation}), we saw that early on at 1 hr after unfolded protein dose there was a marked increase in ATF4 and CHOP regulation, ATF6 signaling along with unfolded protein sensing and degradation.
These are hallmarks of the adaptation-alarm phase of the UPR response (Fig. \ref{fg:cross_plot_flux}D).
The apoptosis phase ($\geq$ 8-10 hrs P2) was marked by increased BiP regulation, enhanced ATF4 transcriptional activity, increased mitochondrial membrane permeability, and increased apoptotic fluxes resulting in cell commitment to apoptosis mediated cell death.
Certain modules were seen to reduce like IRE1-TRAF2 signaling, ASK1 activation.
However, not much difference was seen in terms of apoptotic fluxes, denoting the commitment of the cell to death.
On the contrary, if we reduced the load of proteins in the adaptation-alarm phase (P4), we saw that the cell could recuperate using its ERAD machinery and the regulation of BiP (Fig. \ref{fg:cross_plot_flux}E-F).

We further investigated the effect of simulated knockout (KO) and overexpression (OX) of key proteins on the UPR system
(supplementary materials Fig. S\ref{fg:Flux_Supp_Fig1}).
% Interestingly, PERK and ATF4 KO studies revealed a slower and lower amount of BiP production ($\sim$ 50\%) as compared to WT. However, ATF6 or IRE1 KO did not affect BiP regulation as compared to WT. This highlighted the dominant role of ATF4 in regulation of BiP, which is consistent with experimental evidence \cite{ma2003delineation}. Regulation of BiP was the critical regulator of spliced XBP1 (XBP1s), which in turn acts as a key marker of progression through different stages of UPR (supplementary materials Fig. S\ref{fg:Flux_Supp_Fig1}E). ATF4, cleaved ATF6, and XBP1s act as integrators of the signals coming from all the three branches of UPR and furthermore leads to regulation of BiP, thereby leading to a negative feedback or control of UPR signal. Another interesting note was the regulation of pro-apoptosis phenotype via regulation of Bcl2. PERK and ATF4 KO led to delay in the onset of apoptosis (marked by slower and lower reduction of Bcl2 levels, supplementary materials Fig. S\ref{fg:Flux_Supp_Fig1}F). This effect could be attributed to the lack of CHOP mediated branch of Bcl2 regulation. On the other hand, IRE1 and CHOP KO leads to drastic reduction in apoptosis (marked by little or no change of Bcl2 levels, supplementary materials Fig. S\ref{fg:Flux_Supp_Fig1}F). CHOP KO implicated the importance of CHOP in the down-regulation of Bcl2. IRE1 KO implicated the critical role of IRE1-TRAF2 mediated route of apoptosis.
Overall signal flow analysis highlighted the extensive amount of crosstalk within the three branches of the UPR network.
Taken together, the population of UPR models recapitulated both short- and long timescale behavior following overload of unfolded proteins in a variety of cell types.
It also captured the integration between multiple pathways and generated specific and testable hypothesis about the role of network components in signal propagation.

>>>>

% which was composed of a system of ordinary differential equations.
% The objective of this study was to assemble a series of molecular modules describing different aspects of UPR and subsequently analyze these models for fragility and robustness.
% Using POETs and a cross-validation scheme, we developed an ensemble of models, consistent with literature data.
%Cellular stressors routinely lead to malfunctions in the folding state of critical proteins, thereby playing a vital role in disease states like cancer, diabetes and cardiovascular ailments. Newly synthesized proteins must fold and assemble into unique three-dimensional structures, in order to become functionally active.

\begin{abstract}
Cells monitor protein folding by an inbuilt quality-control system in which incorrectly or misfolded folded proteins are tagged for degradation or sent back through a refolding cycle.
However, continued accumulation of incorrectly folded proteins triggers the Unfolded Protein Response (UPR), which attempts to re-establish folding homeostasis or commits the cell to apoptosis.
In this study, we developed a family of mechanistic models of the mammalian UPR system.
An ensemble of models parameters was estimated by minimizing the difference between simulations and experimental measurements using multiobjective optimization.
The ensemble of model parameters was validated using cross-validation.
Analysis of the model ensemble suggested the three branches of UPR fired simultaneously where the counteracting effects of the ER stress transducers converge at feedback regulation of the molecular chaperone BiP.

Model analysis highlighted the presence of a sequential order for firing of the ER stress transducers in UPR.
The three main regulators of feedback, ATF4, cleaved ATF6 and XBP1s share the load ensuring the phased response of UPR.
However these regulators add on to the overall fragility of the system and allow scope for manipulation of UPR as seen by sensitivity/robustness analysis.
Downstream effects of UPR include a balance between cell survival and cell death based on the magnitude of the stress.
Model analysis suggested that the cell-death axis was relatively robust owing to redundant routes e.g., APAF-1 dependent and APAF-1 independent routes of apoptosis.
However the cell-survival axis was relatively susceptible to perturbations owing to multiple levels of regulation, thereby highlighting importance of key proteins like CHOP.
\end{abstract}
\begin{document}

@@

\subsection*{Formulation of the UPR network architecture}
The UPR network described the ER folding cycle, ER-associated degradation (ERAD), ER-stress transducer (PERK, IRE1 and ATF6) signaling cascades and stress-induced caspase activation (Fig. \ref{fg:fig1}).
The network consisted of 636 protein or mRNA species interconnected by 1090 interactions (Fig. \ref{fg:fig1} Inset).
Connectivity was formulated from a comprehensive review of the primary literature  \cite{naidoo2009er, ron2002translational, kaufman2002unfolded,ellgaard2003qce,Fonseca:2009fk,Schroder:2005vn,gotoh:hdc,mccullough2001gsc, belmont2008cga,Hetz:2009qc,Urano:2000uq,szegezdi2006mediators}, and from on-line databases; String-8 \cite{Jensen:2009fk}, NetworKIN \cite{Linding:2007fk} and TRANSFAC.
Model connectivity was not specific to a single cell-line; rather, it was a canonical representation of the pathways involved in monitoring and controlling the folding capacity of a generic well-mixed ER compartment.
UPR induction was modeled as the release of BiP from the ER stress transducers, PERK, IRE1$\alpha$, and ATF6 leading initially to adaptation of the folding cycle and then, subsequently, to alarm and apoptosis.
The adaption phase of UPR was marked by general translation attenuation, selective transcriptional programs for key species like bZIP transcription factor ATF4 \cite{lu2004tra}, cellular inhibitor of apoptosis (cIAP) \cite{hamanaka2008pdr}, molecular chaperones e.g., BiP \cite{harding2003isr}, and enhanced clearance of accumulated proteins via ERAD.
The alarm and apoptosis phases were mediated by the induction of CHOP \cite{ron1992cnd}, regulation of Bcl2, Bcl2-antagonist of cell death (BAD) \cite {wang1999cia} and (TNF) receptor associated factor 2 (TRAF2) \cite{lei2003jpb,putcha2003jmb, yamamoto1999bpa, szegezdi2006mediators} activation (Fig. \ref{fg:fig1}).
Model connectivity is available from the GitHub model repository, and is described further in the supplementary materials.

%\clearpage
% \subsection*{Estimating a population of canonical UPR models using POETs.}

% Despite a significant identifiability challenge, the multiobjective POETs algorithm generated a predictive UPR model population.
% The three phases of UPR were modeled using mass action kinetics within an ordinary differential equation (ODE) framework.
% While ODEs and mass-action kinetics are common methods of modeling biological pathways \cite{Schoeberl:2002rv, Chen:2009uq, Helmy:2009kx}, this modeling strategy resulted in a large number of unknown model parameters.
% These parameters were not uniquely identifiable (data not shown).
% Instead, we estimated an experimentally constrained population of parameters using multiobjective optimization.

\subsection*{The UPR model recapitulated the adaptation, alarm, and apoptotic phases of UPR}
A population of unknown model parameters was estimated from 33 dynamic and steady state data sets taken from literature (Table \ref{obj-table}).
The residual between model simulations and each of the measurement sets was simultaneously minimized using the multiobjective POETs algorithm \cite{Song:2010ij}.
A leave-eight-out cross-validation strategy was used to independently estimate the training and prediction error over the 33 data sets; we estimated four different model families,
where eight of the 33 objectives were reserved for validation and 25 were used for model training.
Starting from an initial best-fit parameter set (nominal set), more than 25,000 probable models were estimated by POETs from which we selected N = 100 models (25 from each training family) with a Pareto rank of one or less (from approximately 1200 possible choices).
The nominal, training (75 models), and prediction (25 models) errors were calculated for each objective (Table \ref{obj-table}).
Models used for prediction error calculations for a particular objective were \emph{not} trained on that objective.
The prediction likelihood was statistically significantly better for 31 of the 33 objective functions at a 95\% confidence level, compared with random parameter sets generated from the nominal set (Table \ref{obj-table}).
POETs generated model families that predicted approximately 94\% of the training data with a significantly higher likelihood than a random control.
However, the specific value of any given parameter was likely not well described; the coefficient of variation (CV) for the model parameters ranged from 0.5 - 1.6, where approximately 65\% of the parameters were constrained with a CV $\leq$ 1.0 (supplementary materials Fig. \ref{fg:Para_S1}).
The most constrained parameters involved a wide-array of functions e.g., regulation of PERK, eIF2$\alpha$, ATF4, Calcineurin, BiP, CHOP and ATF6 signaling.
However, the least constrained parameters involved JNK and apoptosis interactions.
POETs identified Pareto fronts between several objectives, e.g., O13$\times$O14, O25$\times$O29, O11$\times$O29, and O27$\times$O2, in the training data  (Fig. \ref{fg:objective_plot}).
Strong Pareto fronts suggested an inability to simultaneously model different aspects of the training data.
However, fronts could also result from experimental artifacts, e.g., variation between cell-lines, time-scale differences, or from functional relationships in the data.
Globally, adaptation and alarm phase training constraints conflicted with those involving apoptosis.
For example, objectives involving caspase-7 or caspase-9 activity conflicted with phosphorylated eIF2$\alpha$ levels.
Phosphorylation of eIF2$\alpha$ by activated PERK attenuates translation, which decreases the ER folding load.
Thus, eIF2$\alpha$ phosphorylation is a key early adaptive event in UPR. On the other hand, caspase-9 is a stress-induced death marker activated only after UPR has failed to restore folding homeostasis.
Conflicts between these early and late phase markers suggested the UPR time scale was perhaps cell-line or perturbation dependent. Next, we compared model simulations with measurements of components mediating the three phases of UPR.

%For example, XBP1 mRNA measurements (O14) conflicted with CHOP protein measurements (O13), even though these data-sets were taken from the same study and were collected in the same cell-line. XBP1 splicing increased BiP levels, which in turn reduced CHOP protein levels, hence the trade-off. Lastly, in addition to fronts, we also observed strong correlation between objectives. For example, models that performed well for the CHOP protein (O11), also performed well against Procaspase-12 (O22) measurements, even though these were not in the same cell-line or from the same study. Both CHOP and Procaspase-12 are downstream of the IRE1/TRAF2/JNK signaling cascade, so these errors were directly correlated.
%\clearpage
% \subsection*{The population of UPR models recapitulated adaptation, alarm, and apoptotic events across multiple cell-lines and timescales.}
%The counteracting effects are further substantiated in simulated knockout studies, wherein knockout of one ER stress transducer led to enhancement of the other branches of UPR (supplementary materials Fig. \ref{fg:Flux_Supp_Fig1}A-C).
%We compared the ability of the system to recuperate upon reversal of unfolded protein loads from the alarm phase (P4, Fig. \ref{fg:proof_of_concept_simulation} ) and the apoptosis phase (P2, Fig. \ref{fg:proof_of_concept_simulation} ). If we reduced the protein load after the cell was committed to apoptosis (as in P3), we found that the cell continued to function similar to the UPR state even upon protein load reduction after 25 hrs (Fig. \ref{fg:cross_plot_flux}C).

We estimated model parameters using measurements from UPR initiation and the downstream activation of apoptosis (Fig. \ref{fg:simulation_comparison}).
We assumed the action of thapsigargin (Tg), a non-competitive inhibitor of SERCA Ca$^{2+}$ transporters, and other stress-inducing agents such as dithiothreitol (DTT) was identical;
All stress agents led to the dissociation of BiP from the ER-stress transducers.
The population of UPR models recapitulated the timescale of PERK phosphorylation (Fig. \ref{fg:simulation_comparison}A) as well as its downstream signaling activity, for example, the phosphorylation of eIF2$\alpha$ (Fig. \ref{fg:simulation_comparison}H).
The nuclear fraction of ATF4 increased from approximately zero (untreated cells) to a maximum value 4 hrs after Tg exposure.
While the model ensemble generally predicted the correct trend, there was significant error in the early time points for ATF4 (Fig. \ref{fg:simulation_comparison}G).
The phosphorylation of eIF2$\alpha$ by PERK is required for ATF4 activation. Interestingly, when we compared model simulations of p-eIF2$\alpha$ levels following Tg (1$\mu$M) exposure in mouse embryonic fibroblasts (MEFs) with measurements (O15 \cite{yamamoto2007tim}), the model correctly captured the appropriate behavior.
To test the functionality of the ATF6 branch of the UPR model, we compared simulations with measurements of cleaved ATF6 in tunicamycin-treated MEFs \cite{lee2002imu}.
ER stress is known to lead to the release of BiP from ATF6.
Cleaved ATF6 is then translocated to the nucleus where it up-regulates gene expression \cite{silver1999mtf,hai1989tfa}.
Simulations of cleaved ATF6 levels following UPR initiation were consistent with measurements (Fig. \ref{fg:simulation_comparison}C).
Signals from the ER stress transducers converge downstream to regulate BiP transcription \cite{malhotra2007era, rao2004mpe,kokame2001iei, yoshida2000aap}.
The ensemble recapitulated the correct trend BiP expression in experiments done on HEK293 cells following stress (Fig. \ref{fg:simulation_comparison}D) \cite{lin2007ire1}.
One of the long-term outcomes of PERK/IRE1 activation is apoptotic cell death.
The link between UPR and apoptosis occurred through the action of eIF2$\alpha$, the dual role of the ATF4 transcription factor and caspase activation by the IRE1-TRAF2 signaling axis.
We constrained model parameters associated with the activation of cell-death using measurements of pro/caspase-7 levels, pro/caspase-9 levels, pro/caspase-3 levels, pro/caspase-12 levels and PARP cleavage mediated by executioner caspases following treatment with 0.5$\mu$M Tg \cite{rao2002cer}.
These experiments were performed in Sak2 cells that lacked Apaf-1 protein expression \cite{rao2002cer}.
Thus, the data allowed us to include a non-Apaf-1 mediated stress-induced caspase activation pathway into the model.
The population of models recapitulated caspase-3 (Fig. \ref{fg:simulation_comparison}K) and caspase-9 (Fig. \ref{fg:simulation_comparison}J), as well as cleaved PARP levels (Fig. \ref{fg:simulation_comparison}L) following exposure to ER stress-inducers.
Interestingly, while PERK activation occurred on the timescale of minutes, initiator and executioner caspase activation occurred over 36 hrs.
Thus, the population of UPR models captured complex signaling events occurring across multiple time scales.
Next, we looked at the signal flow through the three branches following the induction of ER stress.

% Certain modules were seen to reduce like IRE1-TRAF2 signaling, ASK1 activation.
% However, not much difference was seen in terms of apoptotic fluxes, denoting the commitment of the cell to death.

\subsection*{Signal flow analysis suggested the UPR branches fired simultaneously.}
Traditionally, it has been hypothesized that there is a sequential order for firing of the ER stress transducers in UPR.
The PERK and ATF6 branches are thought to be activated before IRE1 \cite{szegezdi2006mediators} and largely promote ER adaptation to misfolding, while IRE1 transmits both survival and pro-apoptotic signals.
However, analysis of the UPR model ensemble suggested the three branches fired simultaneously, and that adaptation, alarm, or apoptosis was the result of counteracting effects of the three UPR signaling pathways (Fig. \ref{fg:proof_of_concept_simulation} and Fig. \ref{fg:Flux_Supp_Fig1}). UPR induction was controlled by manipulation of the generation rate of unfolded or misfolded protein (qP) in the ER compartment.
Upon induction, initially ($t\leq$ 1hr) the response was damped, marking the adaptation phase of UPR (Fig. \ref{fg:proof_of_concept_simulation}A).
Adaptation was followed by increased IRE1$\alpha$, PERK, and ATF6 activity at $t\sim$ 1 hr, marking the onset of the alarm phase.
The alarm phase was followed by a steady state at $t\sim$ 8-10 hrs, marking the onset of the apoptosis phase of UPR (Fig. \ref{fg:proof_of_concept_simulation}A).
Our analysis was substantiated further by looking at the fluxes at different phases of UPR induction (Fig. \ref{fg:proof_of_concept_simulation}B).
At P1, there was a marked increase in ATF4 and CHOP activity, ATF6 signaling and unfolded protein sensing and degradation by ERAD.
These are hallmarks of the adaptation-alarm phase of the UPR response (Fig. \ref{fg:cross_plot_flux}D).
The apoptosis phase at P2 ($t\geq$ 8-10 hrs) was marked by increased BiP regulation, enhanced ATF4 transcriptional activity, increased mitochondrial membrane permeability, and increased apoptotic fluxes.
On the other hand, if we reduced the load of unfolded protein in the adaptation-alarm phase (P4), the cell recuperated using its ERAD machinery and the regulation of BiP (Fig. \ref{fg:cross_plot_flux}E-F).
Thus, following initiation of UPR all three sensor branches were simultaneously activated. However, it was unclear which parameters or species controlled the responses of the three branches.
We addressed this question using sensitivity and robustness analysis.

% We further investigated the effect of simulated knockout (KO) and overexpression (OX) of key proteins on the UPR system
% (supplementary materials Fig. \ref{fg:Flux_Supp_Fig1}).
% Interestingly, PERK and ATF4 KO studies revealed a slower and lower amount of BiP production ($\sim$ 50\%) as compared to WT. However, ATF6 or IRE1 KO did not affect BiP regulation as compared to WT. This highlighted the dominant role of ATF4 in regulation of BiP, which is consistent with experimental evidence \cite{ma2003delineation}. Regulation of BiP was the critical regulator of spliced XBP1 (XBP1s), which in turn acts as a key marker of progression through different stages of UPR (supplementary materials Fig. \ref{fg:Flux_Supp_Fig1}E). ATF4, cleaved ATF6, and XBP1s act as integrators of the signals coming from all the three branches of UPR and furthermore leads to regulation of BiP, thereby leading to a negative feedback or control of UPR signal. Another interesting note was the regulation of pro-apoptosis phenotype via regulation of Bcl2. PERK and ATF4 KO led to delay in the onset of apoptosis (marked by slower and lower reduction of Bcl2 levels, supplementary materials Fig. \ref{fg:Flux_Supp_Fig1}F). This effect could be attributed to the lack of CHOP mediated branch of Bcl2 regulation. On the other hand, IRE1 and CHOP KO leads to drastic reduction in apoptosis (marked by little or no change of Bcl2 levels, supplementary materials Fig. \ref{fg:Flux_Supp_Fig1}F). CHOP KO implicated the importance of CHOP in the down-regulation of Bcl2. IRE1 KO implicated the critical role of IRE1-TRAF2 mediated route of apoptosis.
% Overall signal flow analysis highlighted the extensive amount of crosstalk within the three branches of the UPR network.
% Taken together, the population of UPR models recapitulated both short- and long timescale behavior following overload of unfolded proteins in a variety of cell types.
% It also captured the integration between multiple pathways and generated specific and testable hypothesis about the role of network components in signal propagation.

% Five parameter sets were selected from the parent ensemble based on rightful representation of the diversity for the sensitivity calculations (further details in the supplementary materials).
% We selected only a few diverse parameter sets for the sensitivity analysis, because of the computational cost of computing all 1090 coefficients over several hours of simulated time.
% However, we did select diverse sets as shown by the scatter in the CV values (Fig. \ref{fg:Para_S1}).
% Thus, as expected, components such as RNA polymerase, or caspase activation were globally important irrespective of the folding state of the ER.
% More interesting, however, were coefficients that shifted above or below the 45$^{o}$-line in the presence of UPR.
% These points denote differentially important network components.
% Likewise, 12\% of the species were significantly more important in UPR compared with normal protein loads (data not shown).
%This analysis supports the structural analysis results, wherein the signal integrators like XBP1s and transcriptional regulation of BiP via ATF6 and ATF4 are the key regulators of the phased response of UPR.
% Knockout of any individual feedback branch like that of ATF4, ATF6 and XBP1s, the system overall remains equally robust.
% Interestingly,  However . Overall $\sim$ 54 \% of the parameters were differentially less sensitive upon removal of BiP feedback as compared to WT.
